---
title: "WaterContent, WHC, Bulk Density"
author: "ErinJHanan"
---
This code uses:
1. Bulk soil sample weights (field moist; g)
2. Bulk soil sample volume (cm3)
3. Rock weights (optional, if not using put zeros in column; g)
4. Rock volume (optional if not using put zeros in column; cm3)
5. Subsample wet weights (field moist; does not include filter; g)
6. Saturated subsample + filter weights (g)
7. Soil oven dry + filter weights (g)
8. Average saturated and oven dry filter weights (g)

And will calculate:
1. Water content (WC; g/g dry soil)
2. Fractional WC (g/g wet soil)
3. WHC (g/g dry soil)
4. Bulk density (g/cm3)

The script will output the results as an excel file into the working directory folder with "_results" appended to each file name. If you wish to save to a folder other than the folder where the input files are stored, uncomment the two setwd lines (near beginning and end of the for loop) and put in the appropriate file paths. 

The input files can be excel or csv. Note that excel files cannot run if they are open. Each input file must have these column headers (no caps):
code | bulk_soil_weight | rock_weight | rock_vol | subsample_weight | sat_subsample_weight | dry_subsample_weight

code is the sample id. You will also need to enter the average oven dry filter weight and saturated filter weights at the beginning of the code.

Keep console open to view and respond to possible prompts and checks

Install/load packages
```{r, message=F}
#install.packages("tidyverse")
#install.packages("openxlsx")
#install.packages("readxl")
library(tidyverse)
library(openxlsx)
library(readxl)
```

Set constants
```{r, message=F}
core_volume <- 263.5 #volume of bulk soil core in cm3
filter_weight <- 0.58   #average oven-dry weight of filter used for WC/WHC measurements
sat_filter_weight <- 1.93 #average weight of saturated filter used for WC/WHC measurements
```

Read in file(s), check for missing columns, perform calculations, generate outputs
```{r, message=F}
setwd("My_folder")    #folder for input files
file_names <- list.files()  #pulls all the files in the folder

for (a in file_names) {
  #setwd(file path for input files, same as set above)  #use this line if input and output folders differ
#determine file type and use appropriate read function
  if (grepl(".xlsx", a) == T){   
    input <- read_excel(a, col_names = T)
  } else if (grepl(".csv", a) == T) {
    input <- read.csv(a, header = T)
  } else {
    stop(cat("Something went wrong reading file ", a, " Is it a .csv or .xlsx?"))
  }

#check column names
  for (b in c("code", "bulk_soil_weight", "rock_weight", "rock_vol", "subsample_weight", "sat_subsample_weight", "dry_subsample_weight")){
    if (!(b %in% names(input))){
      stop(paste("column name", b, "not found in file", a))
    }
  }

  calcs <- input %>%
    mutate(
      wc = (subsample_weight - (dry_subsample_weight - filter_weight)) / (dry_subsample_weight - filter_weight),
      fwc = (subsample_weight - (dry_subsample_weight - filter_weight)) / subsample_weight,
      whc = ((sat_subsample_weight - sat_filter_weight) - (dry_subsample_weight - filter_weight))/ (dry_subsample_weight - filter_weight),
      bd = ((bulk_soil_weight - rock_weight) * (1-fwc)) / (core_volume - rock_vol)
    )
    
# Save results
  #setwd("folder for saving output files")  #use this line if input and output folders differ
  WC_FWC_WHC_BD <- calcs %>%
    select(code | wc | fwc | whc | bd) 
  save_name <- sub(".csv", "", a) #remove .csv from file name
  save_name <- sub(".xlsx", "", save_name) #remove .xlsx from file name
  write.xlsx(WC_FWC_WHC_BD, paste0(save_name, "_results.xlsx"))
}
```
