---
title: "Microbial Biomass C and N, Exchangeable organic C & N"
author: "ErinJHanan"
---
This script calculates microbial biomass C, N, biomass C:N and exchangeable organic C 

This script will run through all the files in the designated working directory.
To run specific files, put the file name or a list of file names into "files <-" instead of running list.files().

This code uses 6 constants:
1. Correction factor for microbial C (micC_corr; optional; if not using put 1)
2. Correction factor for microbial N (micN_corr; optional; if not using put 1)
3. C blank correction (average NPOC in unfumigated blanks mg/L)
4. N blank correction (average TN in unfumigated blanks mg/L)
5. C_cfe blank correction (average NPOC in fumigated blanks mg/L)
6. N_cfe blank correction (average NPOC in fumigated blanks mg/L)

And 11 data columns:
1. Code (sample ID)
2. Bulk density (g/cm3)
3. Soil weight extracted for unfumigated samples (field moist; g)
4. Soil weight extracted for fumigated samples (field moist; g)
5. Fractional Water Content (g/g of field moist soil)
6. Volume of K2SO4 extractant (mL; we typically use 40mL)
7. Volume of subsample solution (mL; how much is put in TOC vial)
8. Dilution volume (mL)
9. Total Inorganic N (ug/g dry soil)
10. NPOC in unfumigated sample (from TOC; mg/L)
11. NPOC in fumigated sample (from TOC; mg/L)
12. TN (from TOC; mg/L)

***NOTE: 5-9 above are assumed to be the same for fumigated and unfumigated samples

This script outputs an excel file containing:
1. Sample code 
2. Microbial biomass C (ug/g dry soil)
3. Microbial biomass N (ug/g dry soil) 
4. Microbial biomass C:N 
5. Exchangeable organic C (ug/g dry soil) 

The file will have the same name as the input but adds "_results". 
If you wish to save output to a different folder than the folder where the input files are stored, 
uncomment the two setwd lines (near beginning and end of the for loop) and put in the appropriate 
file paths. By default, it saves to the same folder as the input files.

The input can be csv or xlsx (note: excel files cannot run if they are open) and should be set up as follows:  

code | bulk_density | soil_weight_extracted |  soil_weight_extracted_cfe | fwc | K2SO4_vol | subsample_vol | dilution_vol | TIN | NPOC | NPOC_cfe | TN   

You will also need to enter the correction factors under 'set constants' below. If you do not want to
use correction factors, enter 1 for each

Keep console open to view and respond to possible prompts and checks

Install/load packages
```{r, message=F}
#install.packages("tidyverse")
#install.packages("openxlsx")
#install.packages("readxl")
library(tidyverse)
library(openxlsx)
library(readxl)
```

Set constants
```{r, message=F}
micC_corr <- 0.45 #from Beck et al 1997; based on extraction efficiency; 
  #change to 1 if no correction factor
micN_corr <- 0.54 #from Brookes et al 1985; based on extraction efficiency; 
  #change to 1 if no correction factor
NPOC_blank <- 0.0 # average NPOC in unfumigated blanks mg/L
TN_blank <- 0.0 #average TN in unfumigated blanks mg/L
NPOC_cfe_blank <- 0.0 #average NPOC in fumigated blanks mg/L
TN_cfe_blank <- 0.0 #average NPOC in fumigated blanks mg/L
```

Read in file(s), check for missing columns, perform calculations, generate outputs
```{r, message=F}
setwd("My_folder")    #folder for input files
file_names <- list.files()  #pulls all the files in the folder

for (a in file_names) {
#setwd(file path for input files, same as set above)  #use this line if input and output folders differ
#determine file type and use appropriate read function
  if (grepl(".xlsx", a) == T){   
    input <- read_excel(a, col_names = T)
  } else if (grepl(".csv", a) == T) {
    input <- read.csv(a, header = T)
  } else {
    stop(cat("Something went wrong reading file ", a, " Is it a .csv or .xlsx?"))
  }

#check column names
  for (b in c("code", "bulk_density", "soil_weight_extracted", "soil_weight_extracted_cfe", "fwc", 
    "K2SO4_vol", "subsample_vol", "dilution_vol", "TIN", "NPOC", "NPOC_cfe", "TN")){
      if (!(b %in% names(input))){
        stop(paste("column name", b, "not found in file", a))
    }
  }

  calcs <- input %>%
    #first blank correct and convert NPOC/TN for unfumigated and fumigated samples into ug/g dry soil
    mutate(
      NPOC_soil = (((NPOC - NPOC_blank) * (subsample_vol + dilution_vol)) * (K2SO4_vol / 1000)) / 
        (soil_weight_extracted * (1 - fwc)) *1000,
      NPOC_cfe_soil = (((NPOC_cfe - NPOC_cfe_blank) * (subsample_vol + dilution_vol)) * 
        (K2SO4_vol / 1000)) / (soil_weight_extracted_cfe * (1 - fwc)) *1000,
      TN_soil = (((TN - TN_blank) * (subsample_vol + dilution_vol)) * (K2SO4_vol / 1000)) / 
        (soil_weight_extracted * (1 - fwc)) *1000,
      TN_cfe_soil = (((TN_cfe - TN_cfe_blank) * (subsample_vol + dilution_vol)) * 
        (K2SO4_vol / 1000)) / (soil_weight_extracted_cfe * (1 - fwc)) *1000,
      #then calculate microbial biomass C & N, Exchangeable organic C & N 
      micC = (NPOC_cfe_soil - NPOC_soil) / micC_corr,
      micN = (TN_cfe_soil - TN_soil) / micN_corr,
      micCN = micC/micN,
      EOC = NPOC_soil,
      EON = TN_soil - TIN
    )
    
# Save results
  #setwd("folder for saving output files")  #use this line if input and output folders differ
  Microbial_exchangeable_CN <- calcs %>%
    select(code | micC | micN | micCN | EOC | EON) 
  save_name <- sub(".csv", "", a) #remove .csv from file name
  save_name <- sub(".xlsx", "", save_name) #remove .xlsx from file name
  write.xlsx(Microbial_exchangeable_CN, paste0(save_name, "_results.xlsx"))
}
```
